---
title: "A Matter of Life and Death"
date: today
date-format: long
author: "Professor Craig Furfine"
subtitle: "Based on a lecture from Strategic Financial Planning and Wealth Management, FINC-9xx"
format: 
  revealjs:
    slide-number: true
    chalkboard:
      buttons: true
    incremental: true
    logo: Kellogg_H_RGB.png
    footer: Copyright Craig Furfine 2023  
server: shiny
fontsize: "36px"
---

```{r requires, echo=FALSE}
require(xml2)
require(knitr)
require(DT)
```

```{r mwt-functions, echo=FALSE}
#| context: setup
#| include: false
# These two above lines required to run shiny.

xmltodf <- function(file) {
  # Look up Axis.Y in XML data
  nonannuitant_xml <- xml2::xml_find_all(
    xml2::read_xml(file),
    ".//Axis//Y"
  )
  
  # Create age from "t" attribute and mortality quantile from content
  nonannuitant <- data.frame(
    age = as.integer(xml2::xml_attr(nonannuitant_xml,"t")),
    qx  = xml2::xml_double(nonannuitant_xml)
  )
  
  # Return the data.frame
  return(nonannuitant)
}

# Format numeric cash value as $dollars.cents string
format_cash <- function(num, escape=T) {
  if (escape) {
    the_format <- "\\$%#.2f"
  } else {
    the_format <- "$%#.2f"
  }
  return(sprintf(the_format, num))
}

# Eqn 8.5
fairNSP <- function(q,v,P=1,x=NA,n=NA) {
  # Enforce types
  q <- as.numeric(q)
  v <- as.numeric(v)[1]
  x <- as.integer(x)[1]
  n <- as.integer(n)[1]
  
  # If x is defined, trim (assume vector starts at age zero)
  if (!is.na(x) & x > 1) {
     q <- q[-(1:x)]
  }
  
  # Check if n is undefined
  # If so, set n to length of q
  # If not, remove irrelevant length from q
  if (!is.na(n)) {
    stopifnot(n <= length(q))
    q <- q[1:n]
  } else {
    n <- length(q)
  }
  
  # I think p is just the cumulative product of 1-q?
  # In the table, p starts at 0p = 1. Shift by 1.
  p <- c(
    1,
    cumprod(1-q)[-n]
  )
  
  pv_denom <- (1+v)^(1:n)
  #when calculating the single net premium, the first payout is discounted
  pv_denom2 <- pv_denom/(1+v)
  #when calculating the annual premium, the first payment is not discounted
  
  return(data.frame(
    q = q,
    p = p,
    excomp = P*q*p,
    pv_excomp = (P*q*p)/pv_denom,
    dis_sr = p/pv_denom2
  ))
}

annual_premium <- function(q,v,P=1,x=NA,n=NA) {
  NSPtable <- fairNSP(q,v,P,x,n)
  return(
    sum(NSPtable$pv_excomp) / sum(NSPtable$dis_sr)
  )
}
```

```{r process-data, echo=FALSE}
#| context: data
#| include: false

# file at https://mort.soa.org/data/t3153.xml
nonannuitantmale <- xmltodf("t3153.xml")

# file at https://mort.soa.org/data/t3156.xml
nonannuitantfemale <- xmltodf("t3156.xml")

names(nonannuitantmale) <- c("Age","qm")
names(nonannuitantfemale) <- c("Age","qf")

bothmandf <- merge(nonannuitantmale,nonannuitantfemale, by.x="Age")

bothmandf$Age <- bothmandf$Age - 1
bothmandf$Age <- as.integer(bothmandf$Age)

save(bothmandf,file = "mortality.Rdata")
```


::: {style="text-align: center; margin-top: 6em"}
[Life Insurance]{style="font-size:96px; color:green"}
:::

------------------------------------------------------------------------

## Preliminary finance background (A)

-   How much would you pay today for a [**\$100 bill**]{style="color:green"}?

-   How much would you pay today for a [**\$100 bill**]{style="color:green"} I would deliver to you next year?

-   The [**present value**]{style="color:blue"} of a cash flow, [**F**]{style="color:green"}, received in [**t**]{style="color:brown"} years is the amount of money you need to invest today at a rate [**v**]{style="color:red"}, such that your investment grows to [**F**]{style="color:green"} in [**t**]{style="color:brown"} years.

-   Mathematically, this can be written $$PV =\frac{F}{(1+v)^t}$$

------------------------------------------------------------------------

## Preliminary finance background (B)

-   The [**present value**]{style="color:blue"} of multiple cash flows is the sum of the present values of the individual cash flows.
-   A commonly occuring series of cash flows is a [**growing annuity**]{style="color:green"}.
-   For example, your salary for the first five years after graduating Kellogg might be:

. . .

```{mermaid}
flowchart LR
  $200,000.00 --- $206,000.00 --- $212,180.00 --- $218,545.40  --- $225,101.76
```

------------------------------------------------------------------------

## Preliminary finance background (C)

```{mermaid}
flowchart LR
  $200,000.00 --- $206,000.00 --- $212,180.00 --- $218,545.40  --- $225,101.76
```

-   The present value of this cash flow stream can be calculated from:
    -   initial value, [**F**]{style="color:green"}, \$200,000
    -   growth rate, [**g**]{style="color:orange"}, which in this case is 3%
    -   discount rate, [**v**]{style="color:red"}
    -   number of payments, [**n**]{style="color:brown"}, which in this case is 5

------------------------------------------------------------------------

## Preliminary finance background (D)

-   In particular, the PV of a growing annuity can be written

. . .

$$ PV =F\frac{1-\bigl(\frac{1+g}{1+v}\bigl)^n}{v-g} $$

------------------------------------------------------------------------

## And now, more about life insurance

-   What is life insurance?
-   Why buy life insurance?
-   How much life insurance do you need?
-   What are the different forms of life insurance?
-   How is term life insurance priced?

------------------------------------------------------------------------

## What is life insurance?

-   A financial contract where
    -   a policy [**owner**]{style="color:blue"} pays a [**premium**]{style="color:red"} to an insurance company.
    -   an insurance company pays a [**death** **benefit**]{style="color:green"} to the [**beneficiary**]{style="color:purple"} in case of death of the [**insured**]{style="color:brown"}.

. . .

```{mermaid}
flowchart LR
  A[Premium] --> B[Insurance Company]
  B --> C{Does the insured die?}
  C --> |No| D[No payment made]
  C --> |Yes| E[Benefit paid to beneficiary]
```

------------------------------------------------------------------------

## Why buy life insurance?

-   To protect against financial problems associated with the insured's premature death

    -   [**Loss of the insured's future earnings**]{style="color:red"}
    -   Pay for funeral expenses, unpaid medical bills, outstanding debts

-   [**Tax preferred savings**]{style="color:green"} (see later lectures)

## How much life insurance do you need?

-   Answer 1: The Human Capital Approach
-   Calculate the after-tax value of the insured's future income less the insured's future implicit liabilities.

------------------------------------------------------------------------

## How much life insurance do you need?

::: {style="font-size: 0.7em; margin-top: 25px;"}
Example: You turned 30 years old today. Your next pre-tax annual salary (payable at the end of this year) is \$100,000. It is growing at a real rate of 0.5% per year. You plan to work until you are 65 years old. You are married and have two children. You expect to live until 85 years old, and you estimate that you currently have an implicit liability of \$24,000, which is constant in real terms. Assume that the after-tax real discount rate is 2% per year and that your average tax rate will be 25% over your working life. How much life insurance do you need?
:::

. . .

::: {style="font-size: 0.6em;"}
$$ PV_{income} =F\frac{1-\bigl(\frac{1+g}{1+v}\bigl)^n}{v-g}=\$75{,}000\frac{1-\bigl(\frac{1+.005}{1+.02}\bigl)^{65-30}}{.02-.005}=\$2{,}023{,}018$$
:::

. . .

::: {style="font-size: 0.6em;"}
$$ PV_{liabilities} =\$24{,}000\frac{1-\bigl(\frac{1}{1+.02}\bigl)^{85-30}}{.02}=\$ 796{,}194 $$
:::

. . .

::: {style="font-size: 0.7em;"}
Therefore, you need \$2,023,018 - \$796,194 = \$1,226,824 of life insurance.
:::

---

## How much life insurance do you need?

-   Problems with the human capital approach
    -   Sensitive to assumptions on [**salary growth**]{style="color:orange"} and [**valuation rate**]{style="color:red"}
    -   Ignores any current wealth



-   Answer 2: The Needs Approach
    -   Calculate the difference between your [**current financial wealth**]{style="color:green"} (from your personal balance sheet) and the [**present value of your dependents' net cash flow**]{style="color:brown"} (from the personal income statement of your dependents).
-   Note: Both methods imply that your [**need**]{style="color:purple"} for life insurance likely diminishes through time.

------------------------------------------------------------------------

## Types of life insurance

-   [**Term**]{style="color:purple"} life insurance provides protection for a specific term and has no savings components.
-   [**Cash value**]{style="color:purple"} life insurance has a savings component added to the underlying insurance component.

------------------------------------------------------------------------

## Pricing of term life insurance

-   To begin to understand life insurance pricing, we have to understand [**mortality risk**]{style="color:red"}, which is the risk that the insured will die within the term of the policy.

-   To illustrate, consider this mortality table, which I downloaded from the [Society of Actuaries](https://mort.soa.org){preview-link="true"} website.

## A sample mortality table

::: {style="margin-left: 200px;font-size: 15pt;"}
```{r, echo=FALSE}
DT::datatable(
  bothmandf,
  caption = 'Mortality Probability of Healthy Non-Annuitants',
  rownames = FALSE , width = "600px", 
  class = 'compact stripe',
  colnames = c('Age','q (male)', 'q (female)'), 
  filter = 'top', 
  options = list(pageLength = 10, 
                 dom = 'tipr'
            )
) 
```

:::

------------------------------------------------------------------------

## Some technical definitions

-   Define $q_x$ to be the probability that an x-year-old person will die within one year.
-   For example, from the mortality table, you can see that $q_{30}$ for a female is `r sprintf(bothmandf[31,3], fmt = '%#.6f')`.
-   Analogously, define $p_x$ to be the probability that an x-year-old person will survive one year. Note: $p_x=1-q_x$.
-   In this example $p_{30}$ for a female is `r sprintf(1-bothmandf[31,3], fmt = '%#.6f')`.

------------------------------------------------------------------------

## Pricing term life insurance

::: {style="font-size: 0.7em; "}
Example: What is the actuarily fair premium for a 1-year term policy on a 30-year-old healthy female that pays the policy beneficiary $\$1{,}000{,}000$ in event of death? Assume the appropriate valuation rate is 3%.
:::

-   With probability $q_{30}$=`r sprintf(bothmandf[31,3], fmt = '%#.6f')`, the insurance company will need to pay out $1,000,000.
-   Therefore, its expected cash outflow is $$q_{30}*\$1{,}000{,}000 = `r format_cash(bothmandf[31,3]*1000000)`$$
-   Assuming the payout is at the end of the year, the insurance company would need to collect today, at age 30: $$\frac{`r format_cash(bothmandf[31,3]*1000000)`}{1+3\%}=`r format_cash(bothmandf[31,3]*1000000/(1.03))`$$

------------------------------------------------------------------------

## Pricing term life insurance

::: {style="font-size: 0.7em;"}
Example: What is the actuarily fair premium for a 2-year term policy on a 30-year-old healthy female that pays the policy beneficiary $\$1{,}000{,}000$ in event of death? Assume the appropriate valuation rate is 3%.
:::

-   The premium must include the expected cash outflow from the insured surviving the first year and then dying during the second.

. . .

```{=tex}
\begin{align*}
\$1{,}000{,}000 \, p_{30}q_{31} &= \$1{,}000{,}000 \, (`r sprintf(1-bothmandf[31,3], fmt = '%#.6f')`)(`r sprintf(bothmandf[32,3], fmt = '%#.6f')`) \\
& =`r format_cash(1000000*(1-bothmandf[31,3])*bothmandf[32,3])`
\end{align*}
```


-   Therefore the single premium needed for a 2-year term policy is $$`r format_cash(bothmandf[31,3]*1000000/(1.03))`+\frac{`r format_cash(1000000*(1-bothmandf[31,3])*bothmandf[32,3])`}{(1+3\%)^2} = `r format_cash(bothmandf[31,3]*1000000/(1.03)+1000000*(1-bothmandf[31,3])*bothmandf[32,3]/(1.03)^2)`$$



---

## Pricing for longer terms

::: {style="font-size:1.2rem"}
```{r}
# Get pricing table and nsp
pricing_table_f <- fairNSP(q = bothmandf$qf, v = 0.03, P = 1000000, x = 30, n = 10)
nsp_f <- format_cash(sum(pricing_table_f$pv_excomp))

# Add Age and j col
pricing_table_f <- cbind(30:39, 0:9, pricing_table_f[c("q","p","excomp","pv_excomp")])

# Format last two columns
pricing_table_f$excomp <- format_cash(pricing_table_f$excomp)
pricing_table_f$pv_excomp <- format_cash(pricing_table_f$pv_excomp)

knitr::kable(
  pricing_table_f,
  digits = c(0,0,6,6,2,2),
  caption = '10-year term life instuance, $\\$1{,}000{,}000$ for 30-year-old woman',
  row.names = FALSE,
  col.names = c('Age','$j$','$q_{30+j}$', '$_jp_{30}$', 'Expected comp', 'Expected PV'),
  escape = FALSE
)
```
:::

Actuarially fair net single premium is sum of expected PV, `r nsp_f`.

---


## Pricing term life insurance

-   Returning to the 2-year example, let's convert the single premium into annual payments.
-   That is, we need to calculate the annual premium $m$ such that $$`r format_cash(bothmandf[31,3]*1000000/(1.03)+1000000*(1-bothmandf[31,3])*bothmandf[32,3]/(1.03)^2)`=m+m\frac{p_{30}}{1+v}$$
-   Therefore, the annual premium the insurance company charges is $$m=\frac{`r format_cash(bothmandf[31,3]*1000000/(1.03)+1000000*(1-bothmandf[31,3])*bothmandf[32,3]/(1.03)^2)`}{1+\frac{p_{30}}{1+v}}=`r format_cash((bothmandf[31,3]*1000000/(1.03)+1000000*(1-bothmandf[31,3])*bothmandf[32,3]/(1.03)^2)/(1+(1-bothmandf[31,3])/(1.03)))`$$



---


## Annual Premium per \$1,000,000

::: {style="text-align: center; margin-top: 3em"}

```{r}
#| panel: sidebar
shiny::selectInput('sex', 'Sex', list(female = "qf", male = "qm"), "female")
shiny::sliderInput('age', 'Age', value = 30, min = 0, max = 118, step = 1)
shiny::sliderInput('term', 'Term', value = 10, min = 1, max = 30, step = 1)
```

```{r}
#| panel: fill
shiny::tableOutput('plot1')
```

```{r}
#| context: server

# Term can be any number of remaining years
shiny::observe({
  shiny::updateNumericInput(inputId = 'term', max = 119-input$age)
})

output$plot1 <- shiny::renderText(
  format_cash(annual_premium(
    bothmandf[[input$sex]],
    v = 0.03,
    P = 1000000,
    x = input$age,
    n = input$term), escape = F)
)
```
:::
---

## Summary

-   In less than an hour, you've learned how an insurance company prices term life.
-   In reality, the prices you are given will differ because
    - The life insurance company is using different mortality statistics.
    - The life insurance company makes a profit, and therefore isn't actuarily fair.
-   Term life is not an investment. Like most (all?) insurance, you hope you do not need it!


---

## Monthly premia

```{r, echo=FALSE}
terms <- c(5,10,20)

# Calculate monthly premium (annual/12) for each term for men
monthly_premia_m <- sapply(terms, function(n){
  return(
    annual_premium(q = bothmandf$qm, v = 0.03, P = 1000000, x = 30, n = n)/12
  )
})

# Calculate monthly premium (annual/12) for each term for women
monthly_premia_f <- sapply(terms, function(n){
  return(
    annual_premium(q = bothmandf$qf, v = 0.03, P = 1000000, x = 30, n = n)/12
  )
})

knitr::kable(
  cbind(terms, format_cash(monthly_premia_m), format_cash(monthly_premia_f)),
  digits = c(0,2,2),
  caption = 'Actuarially fair monthly premia, $\\$1{,}000{,}000$ for 30-year-old',
  row.names = FALSE,
  col.names = c('Term', 'Male', 'Female'),
  escape = FALSE
) 

```
